Metric Spaces with Expensive Distances

### 2019-11-24 paper/draft_socg.tex
Add some text about EMD.

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
We demonstrate experimentally that on a collection
of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

We performed two types of experiments: on points sampled from a low-dimensional Euclidean
space and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

@@

>>>>

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds \cite{althofer1993sparse}.
Our experiments with Earth Mover distance and persistence
diagrams of 3-D shapes show that blind greedy spanner can speed up computations
by a factor of 3, saving hundreds of hours of machine time,
even though the images and diagrams in our data sets are of moderate size.
%of persistence diagrams of moderate size, computing a blind greedy spanner is faster than (approximately)
%computing all $\binom{n}{2}$ distances.
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms gives much smaller
spanners on the tested examples. 
Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size.

@@

We performed three types of experiments: on points sampled from a low-dimensional Euclidean
space, on images with Earth Mover distance, and on persistence diagrams
with $q$-Wasserstein distance $\wsdist{q}$.\footnote{See \cref{sec:appendix_background} for the definition of $\wsdist{q}$.}
Clearly, our model is not meaningful in Euclidean spaces because
distance computations are cheap.
However, in this case we can easily check the performance
of our heuristics on a variety of easily controllable
data sets.  In order to test the \textsc{BlindQuasiSorted} algorithms, we multiply the true distance by a factor from $[1,2]$ chosen uniformly.
By \textit{sparseness} of a graph with $n$
vertices we always mean the number of edges of the graph divided by $\binom{n}{2}$, the number of edges in the complete graph on $n$ edges.

@@

For experiments with images, we used the German traffic sign
dataset and the code for computing the Earth Mover distance
by Gary Dornan.
We randomly selected 250 images from the benchmark.
The first two columns of \cref{tbl:emd_gtsrb}
compare the sparseness of \bgrdy and greedy spanners.
For smaller values of $\eps$,
the \bgrdy spanner contains approximately 3 times more
edges than the greedy spanner, so the ratio is roughly
the same as for persistence diagrams.
The last 3 columns show the advantage
of \bgrdy in terms of the running time.
For $10 \%$ approximation, we can reduce
the running time by more than a factor of $2$,
saving $98$ hours of computation time.
For $20 \%$ approximation,
we speed up the computations by a factor of 3.

@@

\myparagraph{Earth Mover distance}
Let $\mu_P$ and $\mu_Q$
be two probability measures supported
on finite sets $P = \{p_1, \dots, p_n \}$
and $Q = \{ q_1, \dots, q_m \}$.
An $m \times n$ matrix $C$
is called a \textit{transportation plan},
if it satisfies the following three conditions:
\begin{enumerate}
    \item $c_{i,j} \geq 0$ for all entries $c_{i, j}$.
    \item $\sum_{i = 1}^m c_{i, j} = p_j$ for all $j = 1, \dots, n$.
    \item $\sum_{j = 1}^n c_{i, j} = q_i$ for all $i = 1, \dots, m$.
\end{enumerate}
%By definition, the \textit{cost} of transportation plan $C$
Let $\dist(\cdot, \cdot)$ denote the Euclidean
distance in plane.
The earth mover distance between $\mu_P$ and $\mu_Q$
is defined as
\[
    \inf_C \sum_{i, j} c_{i, j} \dist(p_i, q_j)$,
\]
where $\inf$ is taken over all transportation
plans $C$. Intuitively, $c_{i,j}$ is the
mass concentrated at the point $p_i$ that
is sent to the point $q_j$, and the earth mover
distance is the smallest total amount of work
needed to transform the measure $\mu_P$
into the measure $\mu_Q$.

Let $I_1$, $I_2$ be two greyvalue images.
We divide each greyvalue of $I_{1,2}$
by the sum of all greyvalues in the corresponding image.
Thus we can interpret $I_i$
as a probability measure supported on a rectangle
that has one vertex at the origin and
fully lies in the positive quadrant $\{ (x, y) \| x \geq 0, y \geq 0 \}$.
The earth mover distance between images $I_1$ and $I_2$
is, by definition, the earth mover distance between
these probability measures.
